---
title: "Governance Analytics Final Project: Researching the Possible Connection of District Revenues and Graduation Rates"
author: "Alex Woodward"
date: "3/16/2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r echo = FALSE}

```

# Governance Analytics - Professor Magallanes

## **Executive Summary:**

For the purpose of this paper, I aim to determine if revenue sources, particularly local revenue sources, have a sizable effect on district graduation rates. Washington State School districts often struggle to acquire revenue sources that are large enough to provide the teacher salaries and quality of education needed throughout the state. Many districts rely on local levies to fund expenses. These levies can be difficult to pass and have expiration dates. While school districts receive revenues from both the State and Federal Governments, I aim to answer the question of how much of a difference local revenue sources  have on the success of students. If local revenue sources are shown to have a large effect on success, additional data may prove useful in attempting to develop more permanent local revenue sources for public education.

Through my analysis, I hope to determine if there is a connection between Washington State School District graduation rates, school district revenue from state sources, school district revenue from local sources, and total school district revenue. Using statistical analysis through R, I will determine if there is a correlation between the aforementioned variables and will assess which school districts in Washington State have the highest rates of graduation. My analysis focuses on public school enrollment. 

## **Getting the Data**
For this research, I look at three forms of revenue: state, local and total revenue sources. The revenue information for this revenue source was taken from the US Census Annual Survey data. This survey was the 2015 Annual Survey of School System Finances, authorized by law under Title 13, United States Code, Sections 161 and 182. The most recent year available was 2015. The files were available in .cvs format.

I also used information from the April 2015 Graduation and Dropout Statistics Annual Report from the Washington State Office of Superintendent of Public Instruction in order to determine district level graduation rates.

Lastly, I used the 2015-2016 Enrollment rates from the  Washington State Office of Superintendent of Public Instruction.

```{r cars}


# Libraries
library(readxl)

```

```{r include= FALSE, warnings=FALSE}
library(tinytex)
```





```{r include= FALSE}
temp = tempfile(fileext = ".xlsx") 
dataURL <- "https://github.com/awood59/PROJECT/raw/master/DropOut.xlsx" 
download.file(dataURL, destfile=temp, mode='wb')

# Now we get the data:
DropOut = read_excel(temp, sheet =1)
head(DropOut,2)
```


```{r include= FALSE}
temp = tempfile(fileext = ".xlsx")
dataURL <- "https://github.com/awood59/PROJECT/raw/master/EnrollmentWA.xlsx" 
download.file(dataURL, destfile=temp, mode='wb')
EnrollmentWA = read_excel(temp, sheet =1)
head(EnrollmentWA,2)
```

```{r include= FALSE}
temp = tempfile(fileext = ".xlsx") 
dataURL <- "https://github.com/awood59/PROJECT/raw/master/WAFinance1.xlsx" 
download.file(dataURL, destfile=temp, mode='wb')
WAFinance1 = read_excel(temp, sheet =1)
head(WAFinance1,2)
```


```{r echo= FALSE}
gradenr=merge(EnrollmentWA,DropOut,by.x='District Code',by.y="District Code" )
```

```{r include=FALSE}
DropOutData <- merge(DropOut, EnrollmentWA, by="District Code")
```

```{r include=FALSE}
FullData <- merge(WAFinance1, DropOutData, by="District Name")
```


```{r include=FALSE}
table(is.na(FullData))
```

```{r include=FALSE}
FullData <- na.omit(FullData)
```

```{r include=FALSE}
FullData$`Adjusted 5-Year Cohort Graduation Rate` <- as.numeric(as.character(FullData$`Adjusted 5-Year Cohort Graduation Rate`))
```
```{r include=FALSE}
FullData$`Cohort dropout rate` <- as.numeric(as.character(FullData$`Cohort dropout rate`))
```

Here are descriptive statistics for graduation rate:
```{r}
summary(FullData$`Adjusted 5-Year Cohort Graduation Rate`)
```

And now for Local Revenue:
```{r}
summary(FullData$'Local Revenue')
```

State Revenue:
```{r}
summary(FullData$`State Revenue`)
```
And finally, Total Revenue:
```{r}
summary(FullData$'Total Revenue')
```

Next a look at some statistical information for Local Revene:

Standard deviation:
```{r include=FALSE}
library(DescTools)
```
```{r}
sd(FullData$`Local Revenue`,na.rm = T)
```

Coefficient of Variation
```{r}
sd(FullData$`Local Revenue`,na.rm = T)/mean(FullData$`Local Revenue`,na.rm = T)
```

Skewness
```{r}
Skew(FullData$`Local Revenue`,na.rm = T)
```

Kurtosis
```{r}
Kurt(FullData$`Local Revenue`,na.rm = T)
```

Now let's look for a possible correlation.
```{r include=FALSE}
data=FullData[is.finite(FullData$`Adjusted 5-Year Cohort Graduation Rate`),]
```
```{r fig.width=6, fig.height=4}
hist(FullData$`Adjusted 5-Year Cohort Graduation Rate`)
```
```{r message=FALSE, warning=FALSE}
explanans=names(FullData)[c(3,4,9)]
for (x in explanans){
    par=cbind(FullData[,x],FullData$`Adjusted 5-Year Cohort Graduation Rate`)
    p=cor.test(FullData[,x],FullData$`Adjusted 5-Year Cohort Graduation Rate`)
    print(paste("Pearson:",p$estimate," - Is significative?",p$p.value<0.05))
}
```
```{r message=FALSE, warning=FALSE}
cor(FullData[explanans])
```
```{r include=FALSE}
row.names(FullData)=FullData$`District Name`
```

Let's run a linear regression to see if there are any correlations. We see a correlation between Local Revenue and Graduation Rate is true.
```{r include=FALSE}
LinRegEPI = lm(`Adjusted 5-Year Cohort Graduation Rate` ~ `Total Revenue` + `State Revenue` + `Local Revenue` + `K-12 Total`, 
               data = FullData)
```

```{r message=FALSE, warning=FALSE}
results=coef(summary(LinRegEPI))

data.frame(Coefficient=results[,1],Significant=results[,4]<0.05)

summary(LinRegEPI)$adj.r.squared
```


Next look at scatter plots of the data.
```{r include=FALSE}
var2=data$`State Revenue`

# Saving mean and sd:

mnVar2=mean(var2,na.rm = T)
sdVar2=sd(var2,na.rm = T)


library(ggplot2)
```
```{r fig.width=6, fig.height=4}
#Multivariate Scatter Plot State Revenue.

plot(x=FullData$`Adjusted 5-Year Cohort Graduation Rate`,y=FullData$`State Revenue`) 
```


```{r fig.width=6, fig.height=4}

#Multivariate Scatter Plot Local Revenue.

plot(x=FullData$`Adjusted 5-Year Cohort Graduation Rate`,y=FullData$`Local Revenue`) 

```

```{r fig.width=6, fig.height=4}

#Multivariate Scatter Plot Total Revenue.

plot(x=FullData$`Adjusted 5-Year Cohort Graduation Rate`,y=FullData$`Total Revenue`) 
```



Now plot a histogram of Graduation Rates.
```{r include=FALSE}
data=FullData[is.finite(FullData$`State Revenue`),]
```
```{r include=FALSE}
var=data$`Adjusted 5-Year Cohort Graduation Rate`
```
```{r include=FALSE}
mnVar=mean(var,na.rm = T)
sdVar=sd(var,na.rm = T)
```

```{r message=FALSE, warning=FALSE}
base = ggplot(data, aes(x=var))
hist = base + geom_histogram(fill="blue",
                             color='grey',
                             aes(y=..density..))
```
```{r message=FALSE, warning=FALSE}
histAndNormal = hist + stat_function(fun=dnorm,
                                     color="red",
                                     args=list(mean=mnVar,sd=sdVar))
```
```{r message=FALSE, warning=FALSE}
histAndNormal
```

I want to use a map to visualize the district graduation rates and revenue levels.
```{r include=FALSE}

# Import Map Files

compressedMap= 'https://github.com/awood59/PROJECT/raw/master/WAOFM%20-%20GIS%20-%20School%20District%20Boundary.zip'
```

```{r include=FALSE}
library(utils)
temp=tempfile()
download.file(compressedMap, temp)
unzip(temp)
```

```{r include=FALSE}
(maps=list.files(pattern = 'shp'))
```

```{r include=FALSE}
library(rgdal)
wazipMap <- rgdal::readOGR("geo_export_4f6f751a-05f3-48de-adf9-962964ba024e.shp",stringsAsFactors=F)
```

```{r include=FALSE}
plot(wazipMap,col='green')
```

```{r include=FALSE}
contribWAmap=merge(wazipMap,FullData, by.x='name', by.y='District Name',all.x=F)
```

```{r include=FALSE}
plot(wazipMap,col='white')
plot(contribWAmap,col='light blue',border='grey',add=T)
```

```{r include=FALSE}
library(RColorBrewer)
library(classInt)
```

```{r include=FALSE}
varToPLot=contribWAmap$`Adjusted 5-Year Cohort Graduation Rate`

numberOfClasses = 7
colorForScale='YlGnBu'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPLot, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)
```

```{r include=FALSE}
legendText="Graduation Rates \n (White is Missing)"
shrinkLegend=0.6
title="Graduation Rates Via School District in Washington State 2015"
```
```{r}
plot(wazipMap,col='white',main=title)

# Second change: col = colors[varToPLot]
plot(contribWAmap, col = colorPallette,border='black',add=TRUE)
legend('bottomleft', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```

Now I want to see a map of Local Revenue.
```{r include= FALSE}
varToPLot2=contribWAmap$`Local Revenue`

numberOfClasses = 7
colorForScale='YlOrRd'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPLot2, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)

legendText="Total Revenue \n (White is Missing)"
shrinkLegend=0.6
title="Local Revenue Via School District in Washington State 2015"
```
```{r}
plot(wazipMap,col='white',main=title)
# second change: col = colors[varToPLot]
plot(contribWAmap, col = colorPallette,border='black',add=TRUE)
legend('bottomleft', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```

Lastly, I wanted to see a map of Total Revenue. 
```{r include= FALSE}
varToPLot3=contribWAmap$`Total Revenue`

numberOfClasses = 7
colorForScale='YlGnBu'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPLot3, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)

legendText="Total Revenue \n (White is Missing)"
shrinkLegend=0.6
title="Total Revenue Via School District in Washington State 2015"
```
```{r}
plot(wazipMap,col='white',main=title)

# second change: col = colors[varToPLot]

plot(contribWAmap, col = colorPallette,border='black',add=TRUE)
legend('bottomleft', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```

While this analysis has given us some insight as to whether there is a connection between revenue sources and graduation rates, the next step would be to determine possible correlations between graduation rates and other variables. Some of these variables may include: school size, town income levels, family status, gender, etc. It would be interesting to see what other elements may be effecting graduation rates.


