---
title: "Governance Analytics Final Project"
author: "Alex Woodward"
date: "3/6/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Governance Analytics Final Project - Alex Woodward

## **Executive Summary:**
Here I will add a description for the executive summary. xx

## **Getting the Data**
Explain the data and where it came from.

```{r cars}
temp = tempfile(fileext = ".xlsx") # use always with Excel
dataURL <- "https://github.com/awood59/PROJECT/raw/master/DropOut.xlsx" # This is the link to the first dataset. Washington State School District Graduation and Drop Out Rates
download.file(dataURL, destfile=temp, mode='wb')
# now get it:
DropOut = read_excel(temp, sheet =1)
head(DropOut,2)
```
```{r}
# Now I will retrieve a second data file with information about Washington State Enrollment Statistics.
temp = tempfile(fileext = ".xlsx") # use always with Excel
dataURL <- "https://github.com/awood59/PROJECT/raw/master/EnrollmentWA.xlsx" 
download.file(dataURL, destfile=temp, mode='wb')
# now get it:
EnrollmentWA = read_excel(temp, sheet =1)
head(EnrollmentWA,2)
```
```{r}
# Finally, one last data file. This is a file with financial statistics per school district. I will clean this data as the district names are coded with an attached number, which I need to remove before merging with the other files. In the end, I cleaned the file with Excel, but I will include information for cleaning via R below.
temp = tempfile(fileext = ".xlsx") # use always with Excel
dataURL <- "https://github.com/awood59/PROJECT/raw/master/WAFinance1.xlsx" 
download.file(dataURL, destfile=temp, mode='wb')
# now get it:
WAFinance1 = read_excel(temp, sheet =1)
head(WAFinance1,2)
```

## Data Cleaning
# For the purpose of data cleaning, I first downloaded the orgingal data file.
```{r}
dataLink='https://raw.githubusercontent.com/awood59/data/gh-pages/SchoolFinanceEdit.csv'
```
```{r}
# Next I read the data.
theData=read.csv(dataLink,stringsAsFactors = F,strip.white = T)
```
```{r}
# Then checked the names of the columns.
names(theData)
```
```{r}
table(theData$STATE)
```

```{r}
WAData=theData[theData$STATE=='Washington',]
row.names(WAData)=NULL
```


```{r}
WAData$NAME
```

```{r}
library(stringr)

(WAData$NAMEbad=unlist(lapply(WAData$NAME,word,-1)))

```


```{r}
WAData$NAMEOK=NA
for (pos in 1:nrow(WAData)){
    if(WAData$NAMEbad[pos]!="DISTRICT"){
    cleanValue=sub(WAData$NAMEbad[pos],'',WAData$NAME[pos])
    WAData$NAMEOK[pos]=trimws(cleanValue)
    }else{WAData$NAMEOK[pos]=WAData$NAME[pos]}
}
```

```{r}
table(unlist(lapply(WAData$NAMEOK,word,-1)))
```

```{r}
WAData[word(WAData$NAMEOK,-1)=='DIST',]
```

```{r}
WAData$NAMEbad=unlist(lapply(WAData$NAMEOK,word,-1))
```


```{r}
for (pos in 1:nrow(WAData)){
    if (WAData$NAMEbad[pos]=='DIST'){
        cleanValue=sub("DIST",'DISTRICT',WAData$NAMEOK[pos])
        WAData$NAMEOK[pos]=trimws(cleanValue)
        }
}
```


```{r}
table(unlist(lapply(WAData$NAMEOK,word,-1)))
```



```{r}
(lastBad=names(table(unlist(lapply(WAData$NAMEOK,word,-1))))[-c(4,5)])
```




```{r}
for (pos in 1:nrow(WAData)){
    if (WAData$NAMEbad[pos]%in% lastBad){
        cleanValue=sub(WAData$NAMEbad[pos],'',WAData$NAMEOK[pos])
        WAData$NAMEOK[pos]=trimws(cleanValue)
        }
}

```

```{r}
table(unlist(lapply(WAData$NAMEOK,word,-1)))
```
```{r}
WAData[word(WAData$NAMEOK,-1)=='C',]
```

```{r}
WAData[word(WAData$NAMEOK,-1)=='C',]$NAMEOK='BREMERTON SCHOOL DISTRICT'
```


```{r}
table(unlist(lapply(WAData$NAMEOK,word,-1)))
```

```{r}
length(WAData$NAMEOK)
```

```{r}
length(unique(WAData$NAMEOK))
```

```{r}
names(WAData)
```

```{r, eval=FALSE}
numericColumns =names(WAData)[4:14]
aggregateBY='NAMEOK'
(cleanWA=aggregate(WAData[,numericColumns], by=list(WAData[,aggregateBY]), sum ))


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
