---
title: "Governance Analytics Final Project"
author: "Alex Woodward"
date: "3/12/2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Governance Analytics Final Project - Alex Woodward

## **Executive Summary:**

Here I will add a description for the executive summary. xx

## **Getting the Data**

Explain the data and where it came from.


```{r cars}

# Libraries
library(readxl)

# This is the link to the first dataset. Washington State School District Graduation and Drop Out Rates

temp = tempfile(fileext = ".xlsx") 
dataURL <- "https://github.com/awood59/PROJECT/raw/master/DropOut.xlsx" 
download.file(dataURL, destfile=temp, mode='wb')

# now get it:
DropOut = read_excel(temp, sheet =1)
head(DropOut,2)
```

Now I will retrieve a second data file with information about Washington State Enrollment Statistics.
```{r}
temp = tempfile(fileext = ".xlsx")
dataURL <- "https://github.com/awood59/PROJECT/raw/master/EnrollmentWA.xlsx" 
download.file(dataURL, destfile=temp, mode='wb')
EnrollmentWA = read_excel(temp, sheet =1)
head(EnrollmentWA,2)
```
Finally, one last data file. This is a file with financial statistics per school district. I will clean this data as the district names are coded with an attached number, which I need to remove before merging with the other files. In the end, I cleaned the file with Excel, but I will include information for cleaning via R below.
```{r}
temp = tempfile(fileext = ".xlsx") 
dataURL <- "https://github.com/awood59/PROJECT/raw/master/WAFinance1.xlsx" 
download.file(dataURL, destfile=temp, mode='wb')
WAFinance1 = read_excel(temp, sheet =1)
head(WAFinance1,2)
```

## Data Cleaning
For the purpose of data cleaning, first download the orgingal data file. This data file has a school district column that includes numbers. In order to merge the datasets, the data is cleaned to remove the numbers from the column.

```{r}
dataLink='https://raw.githubusercontent.com/awood59/data/gh-pages/SchoolFinanceEdit.csv'
```
```{r}

# Now read the data.
theData=read.csv(dataLink,stringsAsFactors = F,strip.white = T)
```
```{r}

# Check the names of the columns.
names(theData)
```
Now let's view the data counts per state.
```{r}
table(theData$STATE)
```
Now we remove data that isn't from Washington.
```{r}
WAData=theData[theData$STATE=='Washington',]
row.names(WAData)=NULL
```

Now we can look at the Washington names.
```{r}
WAData$NAME
```

Next remove the unnecessary numbers
```{r}
library(stringr)
(WAData$NAMEbad=unlist(lapply(WAData$NAME,word,-1)))
```
Conintue with the rest of data cleaning.
```{r}
WAData$NAMEOK=NA
for (pos in 1:nrow(WAData)){
    if(WAData$NAMEbad[pos]!="DISTRICT"){
    cleanValue=sub(WAData$NAMEbad[pos],'',WAData$NAME[pos])
    WAData$NAMEOK[pos]=trimws(cleanValue)
    }else{WAData$NAMEOK[pos]=WAData$NAME[pos]}}
```

```{r}
table(unlist(lapply(WAData$NAMEOK,word,-1)))
```

```{r}
WAData[word(WAData$NAMEOK,-1)=='DIST',]
```

```{r}
WAData$NAMEbad=unlist(lapply(WAData$NAMEOK,word,-1))
```

```{r}
for (pos in 1:nrow(WAData)){
    if (WAData$NAMEbad[pos]=='DIST'){
        cleanValue=sub("DIST",'DISTRICT',WAData$NAMEOK[pos])
        WAData$NAMEOK[pos]=trimws(cleanValue)}}
```

```{r}
table(unlist(lapply(WAData$NAMEOK,word,-1)))
```

```{r}
(lastBad=names(table(unlist(lapply(WAData$NAMEOK,word,-1))))[-c(4,5)])
```

```{r}
for (pos in 1:nrow(WAData)){
    if (WAData$NAMEbad[pos]%in% lastBad){
        cleanValue=sub(WAData$NAMEbad[pos],'',WAData$NAMEOK[pos])
        WAData$NAMEOK[pos]=trimws(cleanValue) }}

```

```{r}
table(unlist(lapply(WAData$NAMEOK,word,-1)))
```
```{r}
WAData[word(WAData$NAMEOK,-1)=='C',]
```

```{r}
WAData[word(WAData$NAMEOK,-1)=='C',]$NAMEOK='BREMERTON SCHOOL DISTRICT'
```

```{r}
table(unlist(lapply(WAData$NAMEOK,word,-1)))
```

```{r}
length(WAData$NAMEOK)
```

```{r}
length(unique(WAData$NAMEOK))
```

```{r}
names(WAData)
```

```{r, eval=FALSE}
numericColumns =names(WAData)[4:14]
aggregateBY='NAMEOK'
(cleanWA=aggregate(WAData[,numericColumns], by=list(WAData[,aggregateBY]), sum ))
```

## **Merging the Data**
As the Enrollment and Drop Out Rate datasets both had district codes, merge via "District Code".

```{r}
gradenr=merge(EnrollmentWA,DropOut,by.x='District Code',by.y="District Code" )
```

```{r}
DropOutData <- merge(DropOut, EnrollmentWA, by="District Code")
```

As the Financials and the new datasets both had district codes, merge via "District Name".
```{r}
FullData <- merge(WAFinance1, DropOutData, by="District Name")
```

Now we can take a look at the data frame variable.
```{r}
str(FullData)
```

Now we can look at a summary of the data.
```{r}
summary(FullData)
```

Now we see if the updated data set has any missing values.
```{r}
table(is.na(FullData))
```

Now let's remove NA's.
```{r}
FullData <- na.omit(FullData)
```

Perfect, and now I want to set the graduation and dropout rates to numeric values.
```{r}
FullData$`Adjusted 5-Year Cohort Graduation Rate` <- as.numeric(as.character(FullData$`Adjusted 5-Year Cohort Graduation Rate`))
```
```{r}
FullData$`Cohort dropout rate` <- as.numeric(as.character(FullData$`Cohort dropout rate`))
```

Good, no missing values. Now let's see which districts have the higest and lowest Local, State, and Total Revenues
```{r}
FullData[which.max(FullData$`State Revenue`),] 
```
```{r}
FullData[which.max(FullData$`Total Revenue`),] 
```
```{r}
FullData[which.max(FullData$`Local Revenue`),] 
```
```{r}
FullData[which.min(FullData$`State Revenue`),] 
```
```{r}
FullData[which.min(FullData$`Total Revenue`),] 
```
```{r}
FullData[which.min(FullData$`Local Revenue`),] 
```
We can see 

Ok, now let's look at the districts with the lowest graduation rates.
```{r}
FullData[which.min(FullData$`Adjusted 5-Year Cohort Graduation Rate`),] 
```
```{r}
FullData[order(FullData$`Adjusted 5-Year Cohort Graduation Rate`),]
```

I am currious what the order of School Districts with highest revenue rates looks like. 
```{r}
FullData[order(-FullData$`Total Revenue`),]
```

Now I'd like to see descriptive statistics for total revenue.
```{r}
summary(FullData$'Total Revenue')
```

Standard deviation:
```{r}
library(DescTools)
sd(FullData$`Total Revenue`,na.rm = T)
```

Coefficient of Variation
```{r}
sd(FullData$`Total Revenue`,na.rm = T)/mean(FullData$`Total Revenue`,na.rm = T)
```

Skewness
```{r}
Skew(FullData$`State Revenue`,na.rm = T)
```

Kurtosis
```{r}
Kurt(FullData$`State Revenue`,na.rm = T)
```

Now let's look start creating a correlation.
```{r}
data=FullData[is.finite(FullData$`Adjusted 5-Year Cohort Graduation Rate`),]
```
```{r}
hist(FullData$`Adjusted 5-Year Cohort Graduation Rate`)
```
```{r}
explanans=names(FullData)[c(3,4,9)]
for (x in explanans){
    par=cbind(FullData[,x],FullData$`Adjusted 5-Year Cohort Graduation Rate`)
    p=cor.test(FullData[,x],FullData$`Adjusted 5-Year Cohort Graduation Rate`)
    print(paste("Pearson:",p$estimate," - Is significative?",p$p.value<0.05))
}
```
```{r}
cor(FullData[explanans])
```
```{r}
row.names(FullData)=FullData$`District Name`
```

```{r}
names(FullData)
```

Let's run a linear regression to see if there are any correlations.
```{r}
LinRegEPI = lm(`Adjusted 5-Year Cohort Graduation Rate` ~ `Total Revenue` + `State Revenue` + `Local Revenue` + `K-12 Total`, 
               data = FullData)
```

Now we can view a summary.
```{r}
summary(LinRegEPI)
```

Now we will collect the results of the regression.
```{r}
results=coef(summary(LinRegEPI))

data.frame(Coefficient=results[,1],Significant=results[,4]<0.05)

summary(LinRegEPI)$adj.r.squared
```

Let's develop scatter plots. First, we need to assign variables.
```{r}
var2=data$`State Revenue`

# Saving mean and sd:

mnVar2=mean(var2,na.rm = T)
sdVar2=sd(var2,na.rm = T)

library(ggplot2)

#Multivariate Scatter Plot State Revenue.

plot(x=FullData$`Adjusted 5-Year Cohort Graduation Rate`,y=FullData$`State Revenue`) 
```

```{r}

#Multivariate Scatter Plot Local Revenue.

plot(x=FullData$`Adjusted 5-Year Cohort Graduation Rate`,y=FullData$`Local Revenue`) 

```

```{r}

#Multivariate Scatter Plot Total Revenue.

plot(x=FullData$`Adjusted 5-Year Cohort Graduation Rate`,y=FullData$`Total Revenue`) 
```

Now plot a histogram of Graduation Rates.
```{r}
data=FullData[is.finite(FullData$`State Revenue`),]
```
```{r}
var=data$`Adjusted 5-Year Cohort Graduation Rate`
```
```{r}
mnVar=mean(var,na.rm = T)
sdVar=sd(var,na.rm = T)
```

```{r}
base = ggplot(data, aes(x=var))
hist = base + geom_histogram(fill="blue",
                             color='grey',
                             aes(y=..density..))
```
```{r}
histAndNormal = hist + stat_function(fun=dnorm,
                                     color="red",
                                     args=list(mean=mnVar,sd=sdVar))
```
```{r}
histAndNormal
```

## **Import Washington State School Districts Map**

```{r}

# Import Map Files

compressedMap= 'https://github.com/awood59/PROJECT/raw/master/WAOFM%20-%20GIS%20-%20School%20District%20Boundary.zip'
```

```{r}
library(utils)
temp=tempfile()
download.file(compressedMap, temp)
unzip(temp)
```

```{r}
(maps=list.files(pattern = 'shp'))
```

```{r}
library(rgdal)
wazipMap <- rgdal::readOGR("geo_export_4f6f751a-05f3-48de-adf9-962964ba024e.shp",stringsAsFactors=F)
```

```{r}
plot(wazipMap,col='green')
```

## Next Step
I will follow up by linking the map and plotting graduation rates via district.

```{r}

# I need to see which columns can be connected in the data and the map. We can see we have the column 'name' in the map file.

names(wazipMap)
```
```{r}

# Fortunately, they are the same as the District Name in the Data file.

list(wazipMap@data$name)
```

```{r}

# We can check that to verify.

names(FullData)
```

```{r}

# Let's run one last string to be absolutely sure.

str(wazipMap$name)
```
```{r}
str(FullData$`District Name`)
```

```{r}

## Everything looks good. Now we merge the map file with the data file by District Name. 

contribWAmap=merge(wazipMap,FullData, by.x='name', by.y='District Name',all.x=F)
```

```{r}
# I want to see the number of rows in the new map.
nrow(contribWAmap)
```

```{r}

## Now we plot the map to see that we are representing missing data correctly.

plot(wazipMap,col='white')
plot(contribWAmap,col='light blue',border='grey',add=T)
```

```{r}

## We need to install the correct libraries.

library(RColorBrewer)
library(classInt)
```

```{r}

## Next we assign the variable for plotting.

varToPLot=contribWAmap$`Adjusted 5-Year Cohort Graduation Rate`

numberOfClasses = 7
colorForScale='YlGnBu'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPLot, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)
```

```{r}

## This code is used to assign legend text, colors, and location.

legendText="Graduation Rates \n (White is Missing)"
shrinkLegend=0.6
title="Graduation Rates Via School District in Washington State 2015"
```
```{r}
plot(wazipMap,col='white',main=title)

# Second change: col = colors[varToPLot]
plot(contribWAmap, col = colorPallette,border='black',add=TRUE)
legend('bottomleft', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```
```{r}

# Now I want to see a map of Local Revenue. We create a new variable to plot. I also chose to change the colors.

varToPLot2=contribWAmap$`Local Revenue`

numberOfClasses = 7
colorForScale='YlOrRd'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPLot2, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)

legendText="Total Revenue \n (White is Missing)"
shrinkLegend=0.6
title="Local Revenue Via School District in Washington State 2015"
```
```{r}
plot(wazipMap,col='white',main=title)
# second change: col = colors[varToPLot]
plot(contribWAmap, col = colorPallette,border='black',add=TRUE)
legend('bottomleft', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```
```{r}

## Lastly, I wanted to see a map of Total Revenue. Again, we create a new variable, and rerun the map.

varToPLot3=contribWAmap$`Total Revenue`

numberOfClasses = 7
colorForScale='YlGnBu'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPLot3, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)

legendText="Total Revenue \n (White is Missing)"
shrinkLegend=0.6
title="Total Revenue Via School District in Washington State 2015"
```
```{r}
plot(wazipMap,col='white',main=title)

# second change: col = colors[varToPLot]

plot(contribWAmap, col = colorPallette,border='black',add=TRUE)
legend('bottomleft', legend = names(attr(colorPallette, "table")), 
       fill = attr(colorPallette, "palette"), cex = shrinkLegend, 
       bty = "n",
       title=legendText)
```



